---
title: "Confirmation analysis"
author: "Andrii Zakharchenko"
output: html_document
---

#### Introduction
The aim of this assignment is to get familiar with multivariate confirmation analysis.
As such, the objective of confirmation analysis is to test hypotheses or theories model. These models could be based on a previous empirical research.
For this purpose, you will use multivariate analysis of variance (MANOVA)
to test your hypothesis about given data.

The deadline for this assignment is 4.11.2019.

### Input data
In this tutorial, we will work with a dataset that aims to the quality of potatoes growing in Oregon (for more details see [1]). Each potato is determined
by its size, area, holding temperature, holding period, and cooking method.
Overall quality consists of three aspects: texture score, flavor score, and
moistness score. In this context, the properties of potatoes could be seen as
independent variables; on the other hand, the quality of potatoes are dependent variables. For loading of the dataset, you can use a prepared function
`Load`:

```{r setup, include=TRUE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load the dataset
Load <- function(path = "potato.dat"){
  potatoData <- read.table(path, header = FALSE)
  colnames(potatoData) <- c("Area", "Size" ,"Temp", "Period", "Method", "Texture", "Flavor", "Moistness")
  potatoData$Area <- as.factor(potatoData$Area)
  potatoData$Size <- as.factor(potatoData$Size)
  potatoData$Temp <- as.factor(potatoData$Temp)
  potatoData$Period <- as.factor(potatoData$Period)
  potatoData$Method <- as.factor(potatoData$Method)
  return(potatoData)
}

source("boxTest.R")
library(mvnormtest)
library(MVN)
library(ggplot2)
library(reshape2)
library(ggpubr)

#load the dataset
potatoData <- Load()
potatoMelted <- melt(potatoData, measure.vars=c("Texture", "Moistness", "Flavor"), value.name = "Score", variable.name="ScoreCategory")
```

### MANOVA
There are four main theoretical issues to be considered before running MANOVA.
Not surprisingly, MANOVA has similar assumptions to ANOVA but extended to the multivariable case:

* **Independence**: Observations should be statistically independent.
* **Random sampling**: Data are randomly sampled from the population
of interest.
* **Multivariate normality**: Dependent variables are multivariate normally distributed within each group of the independent variables, which are categorical.
* **Homogeneity of covariance matrices**: The population covariance
matrices of each group are equal.

Take the initial assumptions granted. The assumption of multivariate normality can be tested using R with a test known as the Shapiro test implemented in mshapiro.test in `mvnormtest` package. The assumption of
equality of covariance matrices is often tested using Box’s test that is implemented in `BoxMTest` function in `boxTest.R` file or as a `BoxM` function in biotools package.

### Step by Step
You should go through the following steps:

1. Formalize your MANOVA hypothesis.
Let's say we want to test if potato scores are affected by area.

This means that our $H_0: \mathbf{\mu_1}=\mathbf{\mu_2}$ where $\mu_i$ is mean vector of groups of length 3. We have only 2 grous in Area an 3 independet variables.
Alternative hypothesys is $H_a: \mu_{ik}\neq\mu_{jk}$ for $i\neq j$ and $i$ or $j$ indicate group number.

2. Visualize your data using plot, boxplot, etc. If data have more than two
dimensions project them to lower dimmensionionality. (Optional: use a dimensionality reduction method instead.)
```{r visualization, echo=FALSE}
basePlot <-ggplot(potatoMelted, aes(x=ScoreCategory, y=Score))
area <- basePlot + geom_boxplot(aes(fill=Area))
size <- basePlot + geom_boxplot(aes(fill=Size))
temp <- basePlot + geom_boxplot(aes(fill=Temp))
method <- basePlot + geom_boxplot(aes(fill=Method))
period <- basePlot + geom_boxplot(aes(fill=Period))
ggarrange(area, size, temp, method, period, nrow=3, ncol=2)
```
3. Check the assumptions for MANOVA model.
```{r multivariate normality, echo=FALSE}
responses <- c("Texture", "Moistness", "Flavor")
resp_g1 <- t(as.matrix(potatoData[potatoData$Area==1, responses]))
resp_g2 <- t(as.matrix(potatoData[potatoData$Area==2, responses]))
mshapiro.test(resp_g1)
mshapiro.test(resp_g2)
```
We see that data violets normality assumpsions in second group. Let's have a look at how data are distributed for Area==2

```{r echo=FALSE}
barplot(table(potatoData[potatoData$Area==2,"Texture"]), xlab = "Texture", ylab="Count")
barplot(table(potatoData[potatoData$Area==2,"Flavor"]), xlab = "Flavor", ylab="Count")
barplot(table(potatoData[potatoData$Area==2,"Moistness"]), xlab = "Moistenss", ylab="Count")
```
We see that the data is indeed doesn't resemble normal distribution, however MANOVA is robust against assumptions violation, so we can continue, however we should remove outliers
```{r remove outliers}

```

```{r Homogenity of cov matrices, echo=FALSE}
X<-
cl<-
BoxMTest()
```

4. Use MANOVA to test your hypothesis.
```{r manova, echo=FALSE}

```
5. Discuss the obtained results (the meaning of the individual statistics,
comparison with the visual analysis ad 2, practical implications), compare with the results of simpler MANOVA alternatives (e.g. repeated
ANOVA).

### Submission Form
Submit your solution to the upload system. Submit the directory you have downloaded with only this file modified. 
Write all your code and answers directly into this file and leave the others unmodified.

### References
[1] Mackey, Stockman: Cooking Quality of Oregon-Grown Russet Potatoes,
American Potato Journal, pp. 395–407, 1958.