---
title: "Manova Alloy Example Notebook"
output:
  html_document:
    df_print: paged
---

The example demonstrates that **Manova shows a higher power than a sequence of Anova trials**. The example works with artificial data motivated in mechanical engineering domain.

There is a categorical independent variable "alloy type" and two continuous dependent variables "flexibility" and "strength". Flexibility has a normal distribution and is independent of alloy type. Strength is correlated with flexibility, the alloy C shows an increased strength. Consequently, the null hypothesis that mean flexibility and strength do not change with the alloy type should be rejected in this task.

Manova detects the influence of alloy type on strength earlier than Anova, i.e., it needs fewer observations for weaker dependency.

### Libraries
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(gplots)
library(plyr)
```


### Generate input data 
There are 3 alloy types. We have 30 samples per type. The magnitude of alloy C influence on strength is the main parameter here (the default value is 0.2).
```{r}
alloy<-factor(c(rep("A",30),rep("B",30),rep("C",30)))
flexibility<-rnorm(90,10,1)
effect_size<-0.2
strength<-0.5*flexibility+0.5*rnorm(90,1,1)+effect_size*as.numeric(alloy=="C")
alloys<-data.frame(alloy,flexibility,strength)
```

### Compute some basic statistics
Correlation in groups should be larger than the correlation across groups as it is influenced by the alloy type too.
```{r}
cor.test(flexibility,strength)
cov(alloys[-1])
cor.test(flexibility[alloy=="A"],strength[alloy=="A"])
cor.test(flexibility[alloy=="B"],strength[alloy=="B"])
cor.test(flexibility[alloy=="C"],strength[alloy=="C"])
```

### Plot the data
```{r message=FALSE, warning=FALSE}
plotmeans(flexibility~alloy,alloys,ylim=c(9.2,10.8),ylab="flexibility, strength",yaxt="n",col="blue")
plotmeans((strength+5)~alloy,alloys,add=TRUE,barcol="red",col="red")
legend("topleft",legend=c("flexibility","strength"),col=c("blue","red"),pch=21)

plot(flexibility,strength,col=alloy,pch=19)
abline(lm(strength~flexibility,subset=alloy=="A",data=alloys),col="black",lty=3)
abline(lm(strength~flexibility,subset=alloy=="B",data=alloys),col="red",lty=3)
abline(lm(strength~flexibility,subset=alloy=="C",data=alloys),col="green",lty=3)

legend("topleft",legend=levels(alloys$alloy),col=c("black","red","green"),pch=19)
```

### Run the statistical tests, compare them
Manova tends to reject the null, Anova does not (the actual result depends on the random process of data generation, the parameter setting influnces it as well).
```{r}
summary(aov(flexibility~alloy,alloys))
summary(aov(strength~alloy,alloys))

summary(manova(cbind(flexibility,strength) ~ alloy, data = alloys),test="Wilks")

```

### Calculate the Wilks lambda statistics as in lecture
This piece of code applies the formulas from the lecture. It shows that they match the definition of covariance matrix for T and E. It also demonstrates the way in which lambda get calculated and statistically evaluated.
```{r}
cross_prod_mat<-function(df,vec){cp<-alply(df,1,function(x){t(x-vec)%*%as.matrix(x-vec)});return(Reduce('+',cp))}
cross_prod_mat<-function(df,vec){cp<-alply(df,1,function(x){as.numeric(x-vec)%o%as.numeric(x-vec)});return(Reduce('+',cp))} # the same via outer product %o%
y_grand<-colMeans(alloys[-1])
T<-cross_prod_mat(alloys[-1],y_grand) 
T/(nrow(alloys)-1) # after normalization identical with the covariance matrix above

alloys.g<-split(alloys[-1],alloys$alloy)
y_group<-do.call(cbind,apply(alloys[-1],2,function(x){aggregate(x~alloy,FUN=mean)}))[c(2,4)]
E=mapply(cross_prod_mat,df=alloys.g,vec=as.list(as.data.frame(t(y_group))),SIMPLIFY=F)
E<-Reduce('+',E)

lambda<-det(E)/det(T) 
lambda # matches the manova Wilks stat

bartlett<--(nrow(alloys)-1-(ncol(alloys[-1])+length(levels(alloy)))/2)*log(lambda) # Bartlett's approximation
1-pchisq(bartlett,ncol(alloys[-1])*(length(levels(alloy))-1)) # pvalue based on comparison with chisq quantiles, Manova match again

```
